---
title: "Reply to BMJ 2019;367:l6573"
author: "João Martins, 01.2020"
output:
  tufte::tufte_html: default
  tufte::tufte_handout: default
editor_options: 
  chunk_output_type: console
---

## Background

[BMJ 2019;367:l6573](https://www.bmj.com/content/367/bmj.l6573) reports that female researchers in Life Sciences use fewer positive terms to describe the importance of their research than their male counterparts.  

Below, we investigate whether those observations can be reproduced for grants funded by the Swiss National Science Foundation (SNSF), the leading public research funding organization in Switzerland.

## Main observations  

* In line with the main finding in [BMJ 2019;367:l6573](https://www.bmj.com/content/367/bmj.l6573), *gender* appears to have an effect in the use of positive terms in English grant applications to the SNSF.  However, when accounting for *domain* and *start year* of the grant, the effect of *gender* is less significant.
* With time, the use of positive terms has increased considerably.
* In the domain of *Humanities and Social Sciences*, significant positive terms are significantly less commonly used than in the domain of *Mathematics, Natural- and Engineering Sciences*, or in *Biology and Medicine*.
* As expected, positive terms are more commonly used in grant writing than in paper writing.

## Analysis

For a statistical analysis in `R`, we use the following packages for convenience:

```{r, message=FALSE}
library(here)
library(tidyverse)
library(tidylog)
library(janitor)
library(stopwords)
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
library(broom)
```

The SNSF provides public datasets with information regarding the research domain, the titles, and the abstracts of the grants it funds, as well as the gender of the grantees.

To reproduce the analysis [BMJ 2019;367:l6573](https://www.bmj.com/content/367/bmj.l6573), we will download two datasets available on the SNSF's grant portal at [p3.snf.ch](http://p3.snf.ch/Pages/DataAndDocumentation.aspx):

```{r, eval=FALSE}

download.file(
  "http://p3.snf.ch/P3Export/P3_GrantExport_with_abstracts.csv",
  here("data", "P3_GrantExport_with_abstracts.csv"))

download.file(
  "http://p3.snf.ch/P3Export/P3_PersonExport.csv",
  here("data", "P3_PersonExport.csv"))
```

It's best to read the `.csv` files with the default column type set to `character`:

```{r, message=FALSE, eval=FALSE}
grants <- 
  here("data", "P3_GrantExport_with_abstracts.csv") %>%
  read_csv2(col_types = cols(.default = "c")) %>% 
  clean_names()

grantees <- 
  here("data", "P3_PersonExport.csv") %>%
  read_csv2(col_types = cols(.default = "c")) %>%
  clean_names()
```

Let's cut the `grants` table to the few variables of interest: *title*, *abstract*, *domain*.

```{r, eval=FALSE}
grants <- grants %>%
  select(
    project_number, 
    project_title, 
    project_title_english,
    abstract,
    lay_summary_english,
    discipline_name_hierarchy,
    discipline_name,
    start_date)
```

(As the SNSF accepts grants in several languages, we try to collect as many English abstracts and titles as possible.)  We also need to remove grants for which title and abstract are not available.

```{r, eval=FALSE}
grants <- grants %>%
  mutate(
    project_title = ifelse(
      is.na(project_title_english), 
      project_title,
      project_title_english),
    abstract = ifelse(
      is.na(lay_summary_english), 
      abstract,
      lay_summary_english)) %>%
  select(
    -project_title_english, 
    -lay_summary_english) %>%
  drop_na()
```

We combine the title and the abstract, remove punctuation, double spaces, and parse the text as lists of words.

```{r, eval=FALSE}
grants <- grants %>%
  unite("text", c("project_title", "abstract"), sep = " ") %>%
  mutate(
    text = str_replace_all(text, "[[:punct:]]", " "),
    text = str_to_lower(text),
    text = str_squish(text),
    text = str_split(text, " "))
```

To further tidy the data-set, we separate the column *discipline_name_hierarchy* into *domain* and *subdomain*:

```{r, eval=FALSE}
grants <- grants %>%
  separate(
    discipline_name_hierarchy, 
    c("domain", "subdomain"), ";", fill = "right")
```

We count the use of stop-words for English, German, French, and Italian (heuristic for language detection), and keep the texts where English has the highest score:

```{r, eval=FALSE}

count_stopwords <- function(lang, w) {
  sum(w %in% stopwords(lang))
}

detect_language <- function(words, babel = c("en", "de", "fr", "it")) {
  max <- map_int(babel,count_stopwords, w = words) %>%
    which.max()
  pluck(babel, max)
}

grants <- grants %>% 
  mutate(lang = map_chr(text, detect_language)) %>%
  filter(lang == "en")
```

And we extract the year from the grant start date.

```{r, eval=FALSE}
grants <- grants %>% 
  extract(start_date, "start_year", "([[:digit:]]{4}$)") %>%
  mutate(
    start_year = as.numeric(start_year),
    start_year = cut(start_year, breaks = 5))
```

As for the grantees, we are interested in *gender* and assume only applicants and responsible applicants actually write the grant.

```{r, eval=FALSE}
grantees <- grantees %>%
  select(person_id = person_id_snsf, 
         gender,
         projects_as_responsible_applicant, 
         projects_as_applicant) %>%
  drop_na()
```

We sort the grants according to whether grantees are all female, all male, or mixed.

```{r, eval=FALSE}
df_gender <- grantees %>%
  select(person_id, gender)

df_mapp <- grantees %>%
  select(person_id, project_id = projects_as_responsible_applicant) %>%
  mutate(project_id = str_split(project_id, ";")) %>%
  unnest(project_id) %>%
  semi_join(grants, by = c("project_id" = "project_number"))

df_app <- grantees %>%
  select(person_id, project_id = projects_as_applicant) %>%
  mutate(project_id = str_split(project_id, ";")) %>%
  unnest(project_id) %>%
  semi_join(grants, by = c("project_id" = "project_number"))

df_app <- df_app %>%
  bind_rows(df_mapp) %>%
  distinct(person_id, project_id) %>%
  left_join(df_gender, by = "person_id")

gender_proj <- df_app %>%
  group_by(project_id) %>%
  summarise(
    n_authors = n(),
    share_women = sum(gender == "female") / n_authors) %>% 
  ungroup() %>%
  mutate(share_women = round(share_women, 1))

grants <- grants %>%
  left_join(gender_proj, by = c("project_number" = "project_id")) %>%
  drop_na() %>%
  mutate(
    gender = "mixed",
    gender = ifelse(share_women == 0, "male", gender),
    gender = ifelse(share_women == 1, "female", gender))

```

For consistency with the method presented in [BMJ 2019;367:l6573](https://www.bmj.com/content/367/bmj.l6573), we flag whether positive terms occur at least once:

```{r eval=FALSE}
contains_any <- function(w, terms)
  any(terms %in% w)

terms25 <- c( 
  "novel", "unique", "promising", "favorable", "robust",
  "excellent", "prominent", "supportive", "encouraging", "remarkable",
  "innovative", "unprecedented", "bright", "enormous", "reassuring",
  "creative", "assuring", "hopeful", "astonishing", "spectacular",
  "amazing", "inventive", "phenomenal", "groundbreaking", "inspiring")

grants <- grants %>% 
  mutate(is_positive = map_lgl(text, contains_any, terms = terms25))

grants %>%
  count(start_year, gender) %>%
  kable()

grants %>%
  filter(start_year != "(1997,2002]") %>% # 2 women only
  mutate(
    start_year = fct_drop(start_year),
    start_year = fct_rev(start_year),
    domain = factor(
              domain, 
              c("Biology and Medicine",
                "Mathematics, Natural- and Engineering Sciences",
                "Humanities and Social Sciences"),
              c("LS", "PE", "SH")),
    subdomain = factor(subdomain),
    gender = factor(gender, c("male", "female"), c("M", "F")),
    is_positive = factor(is_positive, c("FALSE", "TRUE"), c("N", "Y"))) %>%
  select(is_positive, gender, 
         domain, subdomain, discipline_name, 
         share_women, start_year) %>%
  write_rds(here("data", "clean_counts.Rds"))
```

And we try to look for possible differences in counts by *gender*, *domain*, and *start year*. Below detailed shares (%):

```{r}
dataset <- read_rds(here("data", "clean_counts.Rds")) %>%
  filter(!is.na(gender))

get_positive_share <- function(grouped_df, side_note) {
  grouped_df %>%
    summarise(
      total  = n(),
      n_pos = sum(is_positive == "Y")) %>%
    mutate(share = round(100 * n_pos / total, 1)) %>%
    kable(caption = side_note)
}

dataset %>%
  group_by(domain) %>%
    get_positive_share(side_note = "Percentage of applications for which title and abstract contain at least one positive term by domain.")

dataset %>%
  group_by(gender) %>%
  get_positive_share(side_note = "Percentage of applications for which title and abstract contain at least one positive term by gender of the grantee.")

dataset %>%
  group_by(start_year, gender) %>%
  get_positive_share(side_note = "Percentage of applications for which title and abstract contain at least one positive term by start year of the grant and by gender of the grantee.")

dataset %>%
  group_by(domain, gender) %>%
  get_positive_share(
    side_note = "Percentage of applications for which title and abstract contain at least one positive term by domain of the grant and by gender of the grantee.")

dataset %>%
  group_by(start_year, domain, gender) %>%
  get_positive_share(
    side_note = "Percentage of applications for which title and abstract contain at least one positive term by start year and domain of the grant and by gender of the grantee.")
```

Generalized linear models (with a logit link function for binary response data) can help checking the significance level of the observed differences:

```{r}
model_gender <- glm(is_positive ~ gender, 
                    dataset, family = "binomial")
model_year <- glm(is_positive ~ start_year, 
                    dataset, family = "binomial")
model_gender_year <- glm(is_positive ~ gender + start_year, 
                    dataset, family = "binomial")
model_gender_year_domain <- glm(is_positive ~ gender + domain + start_year, 
                           dataset, family = "binomial")

tidy(model_gender) %>% 
  kable(caption = "GLM for gender of the grantee.")
tidy(model_year) %>% 
  kable(caption = "GLM for start year of the grant.")
tidy(model_gender_year) %>% 
  kable(caption = "GLM for gender of the grantee and start year of the grant.")
tidy(model_gender_year_domain) %>% 
  kable(caption = "GLM for gender of the grantee and start year and domain of the grant.")
```

